---
title: "Activities - Scatterplot, Correlation, Crime"
author: "Wells"
date: "2025-10-07"
output: html_document

Nia Smith
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Activity #1:

Examine this New York Times scatterplot from 2013:
https://archive.nytimes.com/www.nytimes.com/interactive/2013/05/07/education/college-admissions-gap.html?_r=0

The graphic displays four different scatterplots showing relationships between student demographics and college acceptance rates. For each of the four scatterplots:

a) Describe the relationship (or lack thereof) between the two variables.
b) Note the specific words and phrases the graphic uses to characterize each relationship.

Write 1-2 sentences summarizing each scatterplot.

Name each person in your group who worked on this entry.

**Write Answer #1:**

Based on this scatter plots the green dots represent schools with high graduation rate and students who attend , whereas red dots show higher student attendance, but lower graduation rates. The variables are showing the percentage of freshman in minorities compared to their overall graduation rate. The scatterplots show how minorites attend colleges with lower graduation rates. 

The first graph shows how there are fewer black students at elite colleges that have higher graduation rate. For example UC-San Diego has 1.1% of black students, but a graduation rate of 85.3%.

The second scatter-plot shows how universities with a higher percentage of black students attend schools with lower graduation rates, which are clustered to the left and up. This is a negative correlation.

The third scatter-plot shows how hispanics are more likely to attend community collges, which also has a higher dropout rate (lower graduation rate). 

Specifically for students who are Asian, they are more likely to attend universities that have higher graduation rates, so the graph shows many of the colleges with a lower graduation rate also has a lower percentage of asian students.


-------------------------------------------------------------------------------

Activity #2: Correlation Demonstration

In the following exercise, we will look at data about 16 different consultants, their aptitude scores and their performance scores.

Our question: what is the relationship between a consultant's aptitude and their performance? 

How are they correlated?

Below we calculate a Pearson's correlation coefficient and then create a regression line.

```{r}
library(tidyverse)
```

Load data for aptitude and performance for 16 different consultants
```{r}
aptitude <- c(45, 81, 65, 87, 68, 91, 77, 61, 55, 66, 82, 93, 76, 83, 61, 74)
performance <- c(56, 74, 56, 81, 75, 84, 68, 52, 57, 82, 73, 90, 67, 79, 70, 66)
consultant <- c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16)
#this builds our table
consultants <- data.frame(consultant, aptitude, performance)
consultants
```


### Task 1: Use `summarize()` and the `cor()` function to calculate the correlation between aptitude and performance, using the pearson method.

```{r}
consultants |> summarize(correlation = cor(aptitude, performance, method="pearson"))
```

We see a Pearson's correlation coefficient of 0.783.
What does that mean?

The Pearson correlation scale is from -1 to 1.
-1 is a perfect negative correlation
1 is a perfect positive correlation

How do we interpret the values in between?

-- High Degree: Values between ±0.50 and ±1 suggest a strong correlation.
-- Moderate Degree: Values between ±0.30 and ±0.49 indicate a moderate correlation.
-- Low Degree: Values below +0.29 are considered a weak correlation.
-- No Correlation: A value of zero implies no relationship.

**Question: How would you describe the 0.783 correlation between aptitute and performance?**

**A: Write your answer:**

The correlation of 0.783 shows a stong correlation between aptitude and performance.

### Task 2: Create a scatterplot of aptitude and performance, using `geom_point()`

```{r}

consultants |> 
  ggplot() +
  geom_point(aes(x=aptitude, y=performance)) +
  geom_smooth(aes(x=aptitude, y=performance), method="lm") +
  expand_limits(x = 0, y = 0)

```


The "best fit" line is known as a regression line that helps us understand the relationship between two variables. This predicts the dependent variable based on the independent variable. 

A regression line is a straight line showing a best-fit between independent and dependent variables.

In the example above, aptitude would be the independent variable, or the predictor. That's because we would expect aptitude to predict performance, which is the dependent variable, or the outcome variable.

Independent = predictor
Dependent = outcome

Regression is one step ahead of correlation. It helps predict the values of dependent variables or outcomes with respect to changes in independent or explanatory variables. 

Regression lines help predict outcomes.


-------------------------------------------------------------------------------

Activity #3: 

# PG Murders and Heat

Let's explore the relationship between crime and hot weather using homicide and temperature data from Prince George's County. 

You'll load the data, perform some basic statistical analysis, and create a scatterplot to explore the relationship. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

**write code below**
```{r}
library(tidyverse)
```


### Task 2: Load the dataset from [https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv](https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv)

**write code below**
```{r}
pg_data <- read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv")
```



### Task 3: Use `summarize()` and the `cor()` function to calculate the correlation between murders and temperature, using the pearson method.

**write code below**
```{r}
pg_data |> summarize(correlation = cor(murders, temperature, method = "pearson"))

```



### Task 4: Create a scatterplot of murders and temperature, using `geom_point()`. Your aes() will have temperatures on the x axis, murders on the y axis and add a label for the month (so label=month inside the aes()).

**write code below**
```{r}
pg_data |> 
  ggplot() +
  geom_point(aes(x = temperature, y = murders, label = month)) +
  geom_smooth(aes(x=temperature, y=murders), method="lm") +
labs(title = "Temperature v Murders", x = "temperatures", y = "murders")

```


### Task 5: Add labels to the scatterplot using geom_text(). Inside geom_text() put the same aes() as geom_point in Task 4, then after the aes() add size = 5.

**write code below**
```{r}

pg_data |> 
  ggplot() +
  geom_point(aes(x = temperature, y = murders, label=month)) +
  geom_smooth(aes(x=temperature, y=murders), method="lm") +
  geom_text(aes(x = temperature, y = murders, label = month)) +
labs(title = "Temperature v Murders", x = "temperatures", y = "murders")

```


### Task 6: Describe the results of both the correlation calculation and the scatterplot.

**Write Answer:**


correlation calculation: The relationship between temperature and murders in Prince Georges county is a strong positive correlation
scatterplot: The scatterplot shows that as temperatures rise, murders rise, further showing the positive relationship between the two variables.

When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
