---
title: "Oct 1 Activities: M&Ms; Survey; Sampling"
date: 9-30-2025
---

All three class activities are on this .rmd notebook

Activity #1: Activity: M&Ms - Central Limit Theorem

Working in groups, each group will get a handful of candy from the bag, and then count the number of brwon and all other color candies. Log those in this Google Sheet 

https://docs.google.com/spreadsheets/d/1KOeoVTq2W7L2T9sWjwh5U_2f59GsTeAylVGD53rNWww/edit?usp=sharing

(give your group a nice name).

Take a look at the values that we all get - what do you notice about the patterns? This is an exercise about the central limit theorem. 


The central limit theorem states that the sampling distribution of the mean of any independent, random variable will be normal or nearly normal, if the sample size is large enough.

Then discuss and try to answer these questions:

--How close is your group's mean to the true proportion of 0.5?

--Why do different groups get different results?

--What happens to individual extreme results when we combine all groups?

--How does this relate to polls and surveys in real life?



Write your answer below. And then you may eat the M&Ms.


Looking at the values of I<3 AUDREYHEPBURN and comparing it to the rest of the classes data, there is a common trend of the group having significantly less brown colored M&Ms compared to other colored M&Ms. Most groups percentage sees an above 80% of colored M&Ms.

My groups mean is pretty close to the average of the class. The class average of colored M&Ms is 89.05% while our average was 92.45%

Each group gets different results since each group has a different sampling of M&Ms.

This relates to polling and survey becasue when completing and sending them out it is important to have random sampling to ensure there is fair representation in the results.


**A:**

------------------------------------------------------------------------------

# Activity #2:

Belonging Survey

Working in pairs, take a look back at the main findings from UMD's Belonging Survey:

https://president.umd.edu/internal/belonging


Specifically, focus on the "Sample Description" section and the Appendix. Discuss the differences and between the population and the sample, and what the document says about how the survey attempted to deal with them. Are there any decisions described in the document that raise questions in your minds? Detail the differences you see and your understanding of the methods they employed (it's ok if you don't understand all of them), and add your questions. You should discuss this as a pair but submit individually.

**A:**

The population for this survey was all of the UMD campus (53,455), but the sample was only 23% (12,233 participants) of the whole community. Since certain groups of campus had more representation than others, they collapsed certain categories in order to analyse the data.  

One question I have is how do they determine which categories to collapse and how do they then ensure that it is not causing a significant change in the data.

# TASK 3: SAMPLING

Based on the class sampling demo (available on Elms), sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

**A**
```{r}
library(tidyverse)

```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

**A**
```{r}
set.seed(42)
younguns <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)

head(younguns)

```

### Task 3: Sample 100 rows of data from the dataset
**A**
```{r}
sample_100 <-younguns |> 
  sample_n(100)

```


### Task 4: Calculate the mean age and standard deviation of the sample
**A**
```{r}
sample_100 |> summarize(mean = mean(age), median = median(age), sd = sd(age))

```



```{r}
sample_50 |> summarize(mean = mean(age), median = median(age), sd = sd(age))

```


### Task 5: Calculate the mean age and standard deviation of the population
**A**
```{r}

younguns |> summarize(mean = mean(age), median = median(age), sd = sd(age))
```


### Task 6: Compare the sample and population statistics
**A**

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?


The mean, median, and standard deviation are all pretty similar when comparing the sample to population.

When you do a smaller sample you see a higher standard deviation, but lower mean and median.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
